The Chemistry Development Kit (CDK): atom typing, rendering, molecular formulas, and substructure searching

### 2016-09-29 article.tex

\address[id=um]{
  \orgname{Dept of Bioinformatics - BiGCaT, NUTRIM, Maastricht University}, % university, etc
  %\street{Waterloo Road},
  \postcode{NL-6200 MD},
  \city{Maastricht},
  \cny{The Netherlands}
}
\address[id=nm]{
  \orgname{NextMove Software Ltd},
  %\street{D\"{u}sternbrooker Weg 20},
  \postcode{CB4 0EY},
  \city{Cambridge},
  \cny{UK}
}
\address[id=uppsala]{
  \orgname{Department of Pharmaceutical Biosciences, Uppsala University},
  %\street{D\"{u}sternbrooker Weg 20},
  \postcode{751 24},
  \city{Uppsala},
  \cny{Sweden}
}
\address[id=azg]{
  \orgname{AstraZeneca, Innovative Medicines \& Early Development, Quantitative Biology},
  %\street{},
  %\postcode{LE1 7RH},
  \city{Mölndal},
  \cny{SE}
}
\address[id=jena]{
  \orgname{Chair for Bioinformatics, Friedrich Schiller University},
  %\street{},
  \postcode{07743},
  \city{Jena},
  \cny{Germany}
}
\address[id=idea]{
  \orgname{Ideaconsult Ltd},
  \street{A. Kanchev 4},
  \postcode{1000},
  \city{Sofia},
  \cny{Bulgaria}
}
\address[id=wi_mit]{
  \orgname{Whitehead Institute for Biomedical Research}, 
  \street{9 Cambridge Center},
  \postcode{MA 02142},
  \city{Cambridge},
  \cny{USA}
}
\address[id=miquel]{
  \orgname{Química Clínica Aplicada},
  %\street{4 Hanway Place},
  \postcode{43870},
  \city{Amposta},
  \cny{Spain}
}
\address[id=gilleain]{
  %\orgname{Department of Informatics, University of Leicester},
  \street{4 Hanway Place},
  \postcode{W1T 1HD},
  \city{London},
  \cny{UK}
}
\address[id=nih]{
  \orgname{National Center for Advancing Translational Science},
  \street{9800 Medical Center Drive},
  \postcode{MD 20878},
  \city{Rockville},
  \cny{USA}
}
\address[id=ebi]{
  \orgname{Metabolomics and Molecular Informatics, EMBL-EBI},
  %\street{A. Kanchev 4},
  %\postcode{1000},
  \city{Hinxton},
  \cny{UK}
}
\address[id=leicester]{
  \orgname{Department of Informatics, University of Leicester},
  %\street{},
  %\postcode{LE1 7RH},
  \city{Leicester},
  \cny{UK}
}

@@

While the CDK has purported to be a general purpose cheminformatics
toolkit, older versions were designed by a community with specific
applications in mind, primary among them being structure elucidation. In
addition, an implicit goal of previous versions was to have the CDK
serve as an educational resource to enable students of cheminformatics
to understand the underlying algorithms. This resulted in certain
functionalities, such as fingerprinting~\cite{Clark2014,Cannon2006},
receiving more attention than others, such as stereochemistry. The
outcome was significant variance in performance and features
throughout the toolkit.

The growth of Open Source software over the last 10 years [REF] is evidence
of the ability of communities of developers to develop systems and
processes that lead to high quality software systems for long term
use. The CDK is no different. The adoption of automatic build systems
and quality control methodologies such as unit testing, automated
source code validation, and peer review by fellow developers have
greatly improved the stability of the library, While it has slowed
development somewhat, it has allowed for cleaning up interdependencies
between modules of code, and importantly, has improved the scalability
of the development model.  This has resulted in significant new
functionality in core APIs while maintaining the quality of code
depending on those core APIs.

Examples of new features supported by the improved development model
include InChI functionality~\cite{Spjuth2013}, greatly improved ring
detection algorithms~\cite{May2014}, improvements to the core atom
type perception module that now covers a much more comprehensive set
of elements, charge states and radical species than previous versions,
a more comprehensive fingerprinting API and many speed and stability
improvements.

>>>>

\address[id=um]{
  \orgname{Dept of Bioinformatics - BiGCaT, NUTRIM, Maastricht University}, % university, etc
  %\street{Waterloo Road},
  \postcode{NL-6200 MD},
  \city{Maastricht},
  \cny{The Netherlands}
}
\address[id=nm]{
  \orgname{NextMove Software Ltd},
  %\street{D\"{u}sternbrooker Weg 20},
  \postcode{CB4 0EY},
  \city{Cambridge},
  \cny{UK}
}
\address[id=uppsala]{
  \orgname{Department of Pharmaceutical Biosciences, Uppsala University},
  %\street{D\"{u}sternbrooker Weg 20},
  \postcode{751 24},
  \city{Uppsala},
  \cny{Sweden}
}
\address[id=azg]{
  \orgname{AstraZeneca, Innovative Medicines \& Early Development, Quantitative Biology},
  %\street{},
  %\postcode{LE1 7RH},
  \city{Mölndal},
  \cny{SE}
}
\address[id=jena]{
  \orgname{Chair for Bioinformatics, Friedrich Schiller University},
  %\street{},
  \postcode{07743},
  \city{Jena},
  \cny{Germany}
}
\address[id=idea]{
  \orgname{Ideaconsult Ltd},
  \street{A. Kanchev 4},
  \postcode{1000},
  \city{Sofia},
  \cny{Bulgaria}
}
\address[id=leicester]{
  \orgname{Department of Informatics, University of Leicester},
  %\street{},
  %\postcode{LE1 7RH},
  \city{Leicester},
  \cny{UK}
}
\address[id=wi_mit]{
  \orgname{Whitehead Institute for Biomedical Research}, 
  \street{9 Cambridge Center},
  \postcode{MA 02142},
  \city{Cambridge},
  \cny{USA}
}
\address[id=miquel]{
  \orgname{Química Clínica Aplicada},
  %\street{4 Hanway Place},
  \postcode{43870},
  \city{Amposta},
  \cny{Spain}
}
\address[id=gilleain]{
  %\orgname{Department of Informatics, University of Leicester},
  \street{4 Hanway Place},
  \postcode{W1T 1HD},
  \city{London},
  \cny{UK}
}
\address[id=nih]{
  \orgname{National Center for Advancing Translational Science},
  \street{9800 Medical Center Drive},
  \postcode{MD 20878},
  \city{Rockville},
  \cny{USA}
}
\address[id=ebi]{
  \orgname{Metabolomics and Molecular Informatics, EMBL-EBI},
  %\street{A. Kanchev 4},
  %\postcode{1000},
  \city{Hinxton},
  \cny{UK}
}

@@

While the CDK has purported to be a general purpose cheminformatics
toolkit, older versions were designed by a community with specific
applications, primary among them being structure elucidation. In
addition, an implicit goal of previous versions was to have the CDK
serve as an educational resource to enable students of cheminformatics
to understand the underlying algorithms. This resulted in certain
functionalities, such as fingerprinting~\cite{Clark2014,Cannon2006},
receiving more attention than others, such as stereochemistry. The
outcome was significant variance in performance and features
throughout the toolkit.

Continued community development of the CDK in the last few years, however, has shown that
Open Source software is here to stay. The adoption of automatic build systems and
quality control methodologies such as unit testing, automated source code validation,
and peer review by fellow developers have greatly improved the stability of the library,
At the expense of slowing down the development, it has allowed for cleaning up interdependencies
between modules of code, and has reset the scalability of the development model,
allowing for a lot of new functionality in code APIs while maintaining the quality
of code depending on those core APIs.

This has resulted in important improved functionality, such as a separately published
InChI functionality~\cite{Spjuth2013} and greatly improved ring detection functionality~\cite{May2014},
but also a core atom type perception module, covering a much wider set of elements than
previous CDK versions and capturing many more charged and radical atom species,
a more comprehensive fingerprinting API, many speed and stability improvements, and more.
